# Priori elicitation using meta-analysis

In this chapter, we attempt to replicate the results from Baeten et al. (2013) according to the descriptive information in the paper. We first perform a random-effects meta-analysis of the 8 historical trials for placebo using JAGS to derive the posterior response rate for placebo. We next try to elicit the prior distribution for treatment according to the descriptive information in Baeten et al. (2013)[(Appendix)]. 

<!-- ## Exercise 1 (Individual project (Part 5A)) -->

## Elicitation of the prior distribution for placebo group

**1. Apply the logit-transformation**

$$
X\sim\text{Bin}(n, \pi)
$$

$$
\hat{\phi}_\text{ML}=\text{logit}(\hat{\pi}_\text{ML})
=\log\left(\frac{\hat{\pi}_\text{ML}}{1-\hat{\pi}_\text{ML}} \right)
=\log\left(\frac{x}{n-x} \right)
$$

$$
\text{se}(\hat{\phi}_\text{ML})=\sqrt{\frac{1}{x}+\frac{1}{n-x}}
$$

```{r}
pl_total <- c(107, 44, 51, 39, 139, 20, 78, 35)
pl_case <- c(23, 12, 19, 9, 39, 6, 9, 10)

## Logit-transformation
pl_y <- log(pl_case / (pl_total - pl_case))

## Standard error
pl_se <- sqrt(1/pl_case + 1/(pl_total - pl_case))

prec_s <- 1 / pl_se^2
```

**2. Implement the code for a random effects meta-analysis in JAGS**

```{r}
pl1_modelString <- "model{
  for (i in 1:length(y)) {
    y[i] ~ dnorm(theta[i], prec_s[i]); 
    theta[i] ~ dnorm(mu, prec_tau); 
  }
  
  theta_new ~ dnorm(mu, prec_tau); # predictive distribution for theta
  
  # predictive distribution at the probability scale
  p_new <- exp(theta_new) / (1 + exp(theta_new)); 
  
  mu ~ dnorm(0.0, 1.0E-4); 
  prec_tau ~ dgamma(1.0E-3, 1.0E-3); # just our assumption
}"

writeLines(pl1_modelString, con="MetaAnalysis.txt")
```

Parametrization

$$
\begin{aligned}
  y_i & \sim \text{N}(\theta_i, 1/ \tau^{s}_i) \\
  \theta_i & \sim \text{N}(\mu, 1/\tau) 
\end{aligned}\quad
\text{for }i=1, \cdots, N
$$

$$
\begin{aligned}
  \mu&\sim\text{N}(0,100^2) \\
  \tau&\sim\text{G}(0.001, 0.001)
\end{aligned}\quad
\text{priors for parameters}
$$

Data required for JAGS: 
$$
\begin{aligned}
  y_i &= \log\left(\frac{x_i}{n_i-x_i} \right) \\
  \tau^{s}_i &= 1\Big/\left(\frac{1}{x_i}+\frac{1}{n_i-x_i}\right)
\end{aligned}
$$

```{r, warning=FALSE, message=FALSE}
library(rjags)
library(coda)
library(ggplot2)
```

```{r}
## Generate a data list for JAGS
dat.jags <- list(y=pl_y, prec_s=prec_s)

## Generate a list for initial values for JAGS
## Set initial values and seed for reproducible results 
inits.jags <- list(
  list(mu=100, prec_tau=0.01, .RNG.name="base::Wichmann-Hill", .RNG.seed=314159), 
  list(mu=10, prec_tau=0.05, .RNG.name="base::Marsaglia-Multicarry", .RNG.seed=159314), 
  list(mu=-100, prec_tau=0.001, .RNG.name="base::Super-Duper", .RNG.seed=413159), 
  list(mu=-10, prec_tau=0.1, .RNG.name="base::Mersenne-Twister", .RNG.seed=143915)
)

## Create a JAGS model object
model.jags <- jags.model(
  file = "MetaAnalysis.txt", # text file for JAGS
  data = dat.jags,           # the data list for JAGS
  inits = inits.jags,        # the list for initialization
  n.chains = 4,              # the number of parallel chains for the model
  n.adapt = 4000             # the number of iterations for adaption
)
```

```{r}
## Use the first 4000 iterations as burn-in iterations
update(model.jags, n.iter = 4000)

## Generate posterior samples
fit.jags.coda <- coda.samples(
  model = model.jags,           # JAGS model object
  variable.names = c("p_new"),  # variable to be monitored
  n.iter = 10000,               # number of iterations to monitor
  thin = 1                      # thinning interval for monitors
)
```

**3. Explain what is being done in each line of the code**

See the comments in the code chunks.

**4. Summarize the posterior predictive distribution for p_new**

```{r}
## Print the summary results of the posterior samples generated
summary(fit.jags.coda)
```

```{r, echo=FALSE, include=FALSE}
plot(fit.jags.coda)
```

```{r, fig.show="hold", out.width="50%"}
## Concatenate 4 chains
m.fit.jags.coda <- as.matrix(fit.jags.coda)

d.mcmc <- data.frame(
  Iterations = rep(4001:14000, times=4), 
  ResponseRate = m.fit.jags.coda[, "p_new"], 
  chainType = rep(c("chain1", "chain2", "chain3", "chain4"), each=10000)
)

ggplot(d.mcmc, aes(x=Iterations, y=ResponseRate, color=chainType)) + 
  geom_line(alpha=0.5) + labs(title="Trace of p_new") + theme_minimal() 

ggplot(d.mcmc, aes(x=ResponseRate, y=..density.., fill=chainType, color=chainType)) + 
  geom_density(alpha=0.2) + labs(title="Density of p_new", y="Density") + theme_minimal()

ggplot(d.mcmc, aes(x=1:nrow(d.mcmc), y=ResponseRate, color=chainType)) + 
  geom_line(alpha=0.5) + labs(title="Trace of p_new", x="Index") + theme_minimal() 

ggplot(d.mcmc, aes(x=ResponseRate, y=..density..)) + 
  geom_density(alpha=0.2, color="darkblue", fill="lightblue") + 
  labs(title="Density of p_new", y="Density") + theme_minimal()
```

```{r}
d.summary <- data.frame(t(rbind(
  colMeans(m.fit.jags.coda), 
  apply(m.fit.jags.coda, 2, function(x) sd(x)), 
  apply(m.fit.jags.coda, 2, function(x) var(x)), 
  apply(m.fit.jags.coda, 2, function(x) quantile(x, probs=c(0.025, 0.5, 0.975)))
)))
colnames(d.summary) <- c("Mean", "SD", "Variance", "2.5%", "Median", "97.5%")
knitr::kable(d.summary, digits=5, align="c", caption="Summary statistics")
```

**5. Compare with the results obtained in Baeten's study**

```{r}
d.comparison <- data.frame(
  c(d.summary$Mean, 0.25), 
  c(d.summary$`2.5%`, 0.13), 
  c(d.summary$`97.5%`, 0.4)
)
colnames(d.comparison) <- c("Mean", "2.5%", "97.5%")
rownames(d.comparison) <- c("Meta-analysis", "Baeten's study")
knitr::kable(d.comparison, digits=3, align="c", caption="Comparision")
```

The estimate from the meta-analysis and its 95% credible interval agree with those reported in Baeten et al. (2013)

**6. Apply the function for moment matching**

In Exercise 2 of Worksheet 2, we derived

$$
\begin{cases}
\alpha=\left(\frac{1-\mu}{\sigma^2}-\frac{1}{\mu} \right)\mu^2 \\
\beta=\alpha\left(\frac{1}{\mu}-1 \right)
\end{cases}
$$

```{r}
## Moment matching function
estBetaParams <- function(mean, var) {
  alpha <- ((1 - mean) / var - 1 /mean) * mean ^ 2
  beta <- alpha * (1 / mean - 1)
  return(params = c(alpha=alpha, beta=beta))
}
```

**7. Report the estimates**

```{r}
estBetaParams(mean=d.summary$Mean, var=d.summary$Variance)
```

**8. Explain the approach to a client**

In this exercise, we attempt to replicate the results from Baeten's study and see whether the similar results can be obtained.

* We first gather the data from 8 historical studies 
* We next apply logit-transformation to response rate for placebo group to obtain an approximately normal distribution of logit-transformed rates
* We then conduct a random-effects meta-analysis to derive the posterior predictive distribution of response rate
* We next summarize the posterior predictive distribution of response rates (i.e., mean and 95% credible interval)
* We then apply the function for moment matching to derive the parameters $\alpha$ and $\beta$ of the Beta prior for the response rate in the placebo group.

$~$

<!-- ## Exercise 2 (Individual project (Part 5B)) -->

## Elicitation of the prior distribution for treatment group

```{r}
## Set seed for reproducible results
set.seed(44566)

## Monte Carlo sample size
M <- 100000

## Generate i.i.d. random samples 
p.p <- rbeta(M, shape1=11, shape2=32)  # response rate in Placebo
p.t <- rbeta(M, shape1=0.5, shape2=1)  # response rate in Treatment

## Construct response rate difference (RRD)
rrd <- p.t - p.p
```

```{r, fig.show="hold", out.width="50%"}
ggplot(data.frame(Difference=rrd), aes(x=Difference, y=..density..)) + 
  geom_histogram(position="identity", bins=50, alpha=0.2, color="white", fill=4) + 
  geom_density(alpha=0.2, color=4) + geom_vline(xintercept=0, color=2) + theme_minimal() + 
  labs(title="Histogram/density of the response rate difference", y="Density")

ggplot(with(density(rrd), data.frame(x, y)), aes(x=x, y=y)) + 
  geom_area(aes(x=ifelse(x>0, x, 0), y=y), fill=4, alpha=0.2) + 
  geom_line(color=4) + geom_vline(xintercept=0, color=2) + 
  labs(title="Density plot of response rate difference", x="Difference", y="Density") + 
  geom_text(aes(x=0.3, y=0.3), label="P(RDD>0)", color=4, check_overlap=TRUE) +
  ylim(0, 2) + theme_minimal()
```

```{r, results='hold'}
## Posterior probability of superiority of treatment over placebo
pps <- mean(rrd>0); pps
pps.se <- sqrt(var(rrd>0)/M); pps.se

## 99% confidence interval
c(pps-3*pps.se, pps+3*pps.se)
```

$\text{Beta}(0.5,1)$ is used for treatment group and it has been shown that with such priors there is an approximately 50:50 chance that the response rate on active treatment would be greater than the response rate on placebo.

Approach:

* Generate a vector of i.i.d. random samples from $\text{Beta}(11, 32)$ as the response rates for the placebo group
* Generate a vector of i.i.d. random samples from $\text{Beta}(0.5, 1)$ as the response rates for the treatment group
* Construct the response rate difference between the response rates in the treatment group and the response rates in the placebo group
* Calculate the posterior probability of superiority and the corresponding Monte Carlo standard error
