---
title: "Problemt Set 2"
subtitle: "Global Poverty and Economic Development"
author: "Wenjie Tu"
date: "Fall Semester 2021"
output: pdf_document
---

```{r, echo=F}
# set working directory
setwd("F:/UZH/21Fall/Global-Poverty/PS/PS2")

# set environment language
Sys.setenv(lang = "us_en")

# clear working directory
rm(list=ls())
```

# 1 Theory Question: Intertemporal Choices

Given information:

$$
\begin{cases}
u(c)=\ln(c)\quad\text{utility function} \\
f(x)=\frac{1}{\rho}x^{\rho}\quad\text{production function} \\
\pi_L=1-\pi_H=\frac{1}{2}\quad\text{probabilities of the state of the world} \\
p=1 \quad\text{output price}
\end{cases}
$$

**Perfect Insurance Markets**

## 1.1

Utility maximization:

$$
\begin{aligned}
\max_{b, x_r, x_h}\quad & u(c_1)+\delta\mathbf{E}[u(c_2)] \\
\textrm{s.t.}\quad & 
\begin{cases}
c_1=Y-x_r-x_h-b \\
c_2=R\cdot b+\pi_HA_Hf(x_r)+(1-\pi_H)A_Lf(x_h)
\end{cases}
\end{aligned}
$$

Rewrite maximization:

$$
\mathcal{L}=u(
\underbrace{Y-x_r-x_h-b}_{c_1}
)+
\delta\mathbf{E}[u(
\underbrace{R\cdot b+\pi_HA_Hf(x_r)+(1-\pi_H)A_Lf(x_h)}_{c_2}
)]
$$


FOC w.r.t. $b$:

$$
\frac{\partial\mathcal{L}}{\partial b}=-u'(c_1)+\delta Ru'(c_2)=0
$$

$$
u'(c_1)=\delta Ru'(c_2)\tag{1}
$$

We know that $u(c)=\ln(c)$

$$
\frac{1}{c_1}=\delta R\frac{1}{c_2}\implies
c_2=\delta R c_1\quad\textrm{(Euler Equation)}
$$

## 1.2

FOC w.r.t. $x_r$:

$$
\frac{\partial\mathcal{L}}{\partial x_r}=-u'(c_1)+\delta\pi_HA_Hu'(c_2)
\frac{\partial f(x_r)}{\partial x_r}=0
$$

$$
u'(c_1)=\delta\pi_HA_Hu'(c_2)\frac{\partial f(x_r)}{\partial x_r}\tag{2}
$$

FOC w.r.t. $x_h$:

$$
\frac{\partial\mathcal{L}}{\partial x_h}=-u'(c_1)+\delta(1-\pi_H)A_Lu'(c_2)
\frac{\partial f(x_h)}{\partial x_h}=0
$$

$$
u'(c_1)=\delta(1-\pi_H)A_Lu'(c_2)\frac{\partial f(x_h)}{\partial x_h}\tag{3}
$$

Combine equations (1), (2), (3)

$$
\begin{cases}
u'(c_1)=\delta Ru'(c_2) \\
u'(c_1)=\delta\pi_HA_Hu'(c_2)\frac{\partial f(x_r)}{\partial x_r} \\
u'(c_1)=\delta(1-\pi_H)A_Lu'(c_2)\frac{\partial f(x_h)}{\partial x_h}
\end{cases}
$$


$$
R=\pi_HA_H\frac{\partial f(x_r)}{\partial x_r}=
(1-\pi_H)A_L\frac{\partial f(x_h)}{\partial x_h}\tag{4}
$$



$$
f(x)=\frac{1}{\rho}x^{\rho}\implies
f'(x)=x^{\rho-1}
$$

Rewrite equation (4):

$$
R=\pi_HA_Hx_r^{\rho-1}=(1-\pi_H)A_Lx_h^{\rho-1}
$$

$$
\begin{cases}
x_r=\left(\frac{\pi_HA_H}{R}\right)^{\frac{1}{1-\rho}} \\
x_h=\left(\frac{(1-\pi_H)A_L}{R}\right)^{\frac{1}{1-\rho}}
\end{cases}
$$

With complete markets, the production and the consumption decisions are separable (from equation (4)). Wealth or impatience and other utility parameters do not enter the investment choice.

## 1.3

With Euler Equation $c_2=\delta R c_1$, we can rewrite the second constraint $c_2=R\cdot b+\pi_HA_Hf(x_r)+(1-\pi_H)A_Lf(x_h)$:

$$
\delta R c_1=R\cdot b+\pi_HA_Hf(x_r)+(1-\pi_H)A_Lf(x_h)
$$

Combine with the first constraint:

$$
\begin{cases}
c_1=Y-x_r-x_h-b \\
\delta R c_1=R\cdot b+\pi_HA_Hf(x_r)+(1-\pi_H)A_Lf(x_h)
\end{cases}
$$

All the parameters are given, so we solve two equations with two unknowns ($b$ and $c_1$).

**Incomplete Insurance Markets**

The individual has different consumption in period 2 depending on the state of the world ($H$ or $L$).

$$
\begin{cases}
c_{2H}=R\cdot b+A_Hf(x_r) & \text{with probability } \pi_H \\
c_{2L}=R\cdot b+A_Lf(x_h) & \text{with probability } \pi_L
\end{cases}
$$

## 1.5

Utility maximization:

$$
\begin{aligned}
\max_{b,x_r,x_h}\quad & u(c_1)+\delta\left(
\pi_H\mathbf{E}[u(c_{2H})]+(1-\pi_H)\mathbf{E}[u(c_{2L})]
\right) \\
\textrm{s.t.}\quad & 
\begin{cases}
c_1=Y-x_r-x_h-b \\
c_{2H}=R\cdot b+A_Hf(x_r) \\
c_{2L}=R\cdot b+A_Lf(x_h)
\end{cases}
\end{aligned}
$$


$$
\mathcal{L}=u(
\underbrace{Y-x_r-x_h-b}_{c_1}
)+\delta\left(
\pi_H\mathbf{E}[u(
\underbrace{R\cdot b+A_Hf(x_r)}_{c_{2H}}
)]+(1-\pi_H)\mathbf{E}[u(
\underbrace{R\cdot b+A_Lf(x_h)}_{c_{2L}}
)]
\right)
$$

FOC w.r.t. $b$:

$$
\frac{\partial\mathcal{L}}{\partial b}=-u'(c_1)+
\delta\left(\pi_HRu'(c_{2H})+(1-\pi_H)Ru'(c_{2L})\right)=0
$$

Euler Equation:

$$
u'(c_1)=\delta R\left(\pi_H u'(c_{2H})+(1-\pi_H) u'(c_{2L})\right)\tag{5}
$$

Consumption: marginal utility in period 1 is equal to the expected marginal utility in period 2.

## 1.6

FOC w.r.t. $x_r$:

$$
\frac{\partial\mathcal{L}}{\partial x_r}=
-u'(c_1)+\delta\pi_H A_H\frac{\partial f(x_r)}{\partial x_r}u'(c_{2H})=0
$$

$$
u'(c_1)=\delta\pi_H A_H\frac{\partial f(x_r)}{\partial x_r}u'(c_{2H})\tag{6}
$$

FOC w.r.t. $x_h$:

$$
\frac{\partial\mathcal{L}}{\partial x_h}=
-u'(c_1)+\delta
(1-\pi_H)A_L\frac{\partial f(x_h)}{\partial x_h}u'(c_{2L})=0
$$

$$
u'(c_1)=\delta
(1-\pi_H)A_L\frac{\partial f(x_h)}{\partial x_h}u'(c_{2L})\tag{7}
$$

$$
A_H f(x_r)>A_Lf(x_h)\implies c_{2H}>c_{2L}\implies u'(c_{2H})<u'(c_{2L})
$$

Combine equations (5) and (6):

$$
\begin{cases}
u'(c_1)=\delta R\left(\pi_H u'(c_{2H})+(1-\pi_H) u'(c_{2L})\right) \\
u'(c_1)=\delta\pi_H A_H\frac{\partial f(x_r)}{\partial x_r}u'(c_{2H})
\end{cases}
$$
Investment: underinvestment in risky input

$$
\pi_H A_H\frac{\partial f(x_r)}{\partial x_r}=
R\left(\pi_H+(1-\pi_H)\frac{u'(c_{2L})}{u'(c_{2H})} \right)>R
$$

Combine equations (5) and (7):

$$
\begin{cases}
u'(c_1)=\delta R\left(\pi_H u'(c_{2H})+(1-\pi_H) u'(c_{2L})\right) \\
u'(c_1)=\delta(1-\pi_H)A_L\frac{\partial f(x_h)}{\partial x_h}u'(c_{2L})
\end{cases}
$$

Investment: overinvestment in hedging input

$$
(1-\pi_H)A_L\frac{\partial f(x_h)}{\partial x_h}=
R\left(\pi_H \frac{u'(c_{2H})}{u'(c_{2L})}+(1-\pi_H) \right)<R
$$

Separation fails: investment now depends on wealth and utility parameters.

# 2 Experiments

## 2.1

Research question: the impact of a new type of insecticide-diffuser on malaria rates in a region of Tanzania.

We can run a natural experiment in the region of Tanzania. 

Experimental strategy: randomized controlled trial

* We randomize villages and assign to 50 villages to the treatment group and the other 50 to the control group.
* In the treatment group: households use the new type of insecticide-diffuser.
* In the control group: households do not use the original type of insecticide-diffuser.

Outcome measures:

We can measure the outcome in the following way:

$$
\textrm{Malaria Rate}=\frac{\text{The Number of Infected Individuals in Village i in One Month}}
{\text{Total Population in Village i in One Month}}
$$


* Malaria rate can be measured as the number of infected individuals divided by the population in a village in one month. The incubation period in most cases varies from 7 to 30 days. One month would be a reasonable period to measure the malaria infection rate
* Most malaria cases occur during rainy season. Ideally, we can conduct experiments during summer.


## 2.2

Spillovers:

* Spillovers within the village: there is a positive externality within the village if households use the new type of insecticide-diffuser. Even if some hoseholds in the treatment group do not use the new type of insecticide-diffuser, they can still benefit from the positive externality generated by other households that use the new insecticide-diffuser.

* Geographic spillovers between C and T villages: if the new type of insecticide-diffuser is very effective in repelling mosquitoes, these mosquitoes will migrate to the neighboring control villages, which leads to more intense transmission in villages where the new type of insecticide-diffuser is not used. With this spillover effect, the treatment effect is more likely to be overestimated.

Spillover effect is stronger within the village than across villages.

## 2.3

### 2.3 (a)

I would cluster regressions at a village level to address spillovers within the village.

### 2.3 (b)

I will do a intention-to-treat analysis to test if attrition is a big problem. Households in C villages may use the new type of insecticide-diffuser intentionally or accidentally. I can also instrument the treatment variable with the intent-to-treat variable and estimate the local average treatment effect on compliers. 

\textcolor{cyan}{Collect data on baseline characteristics between groups (those who dropped out and those who stayed), to ensure that there was no selection issues. You would also want to test that your control and treatment groups are still comparable. Would want to ensure that the attrition is random.}

### 2.3 (c)

If the randomization is conducted without any compromise, I would expect estimates to be the same more or less after including baseline covariates. Randomized controlled trials always produce an unbiased estimate for the treatment effect even without adjusting for any baseline covariate. 

# 3 Data Exercise: Difference-in-Differences (DiD)

```{r message=F, warning=F}
library(stargazer)
library(plm)
library(ggplot2)
library(dplyr)
library(did)
```

```{r}
# read in data
d.did <- read.csv("./data_DiD.csv")
```

## 3.1

```{r}
pre.treat <- mean(d.did[d.did$treatment == 1 & d.did$after == 0, "y"])

pre.control <- mean(d.did[d.did$treatment == 0 & d.did$after == 0, "y"])

post.treat <- mean(d.did[d.did$treatment == 1 & d.did$after == 1, "y"])

post.control <- mean(d.did[d.did$treatment == 0 & d.did$after == 1, "y"])
```


```{r results='asis'}
df <- data.frame(
  matrix(data=c(pre.treat, pre.control, post.treat, post.control), 
         ncol=2, nrow=2)
)
rownames(df) <- c("Treatment Group", "Control Group")
colnames(df) <- c("Pre-Program", "Post-Program")

knitr::kable(df)
```


```{r}
cat(sprintf("The DiD estimator is %.3f", 
            (post.treat - pre.treat) - (post.control - pre.control)))
```


## 3.2

```{r, results='asis', message=F, warning=F}
did.ols <- plm(y ~ after*treatment, data=d.did, effect="individual", 
                model="pooling", index=c("town", "t"))
vcov.ols <- vcovHC(did.ols, type="sss", cluster="group")
se.ols <- sqrt(diag(vcov.ols))
stargazer(did.ols, se=list(se.ols), single.row=T, keep.stat=c("rsq", "n"), 
          title="OLS with SE clustered by town", header=F)
```

The estimate from regression is the same as the estimate from the table in (1).

## 3.3

```{r}
cat(sprintf(
"Mean value of x1 in treatment group is %.3f\n
Mean value of x1 in control group is %.3f\n
Mean value of x2 in treatment group is %.3f\n
Mean value of x2 in control group is %.3f\n", 
mean(d.did[d.did$treatment==1, "x1"]), 
mean(d.did[d.did$treatment==0, "x1"]), 
mean(d.did[d.did$treatment==1, "x2"]), 
mean(d.did[d.did$treatment==0, "x2"]))
)
```

<!-- As shown above, the mean value of $x_1$ is quite close between treatment and control towns, while the mean value of $x_2$ is significantly different between treatment and control towns. 

Imbalance between groups in baseline variables that may influence outcome can bias statistical tests (standard errors), which in some sense affects the validity of the DiD approach. -->

The coefficient for $x_1$ is not significant for the first regression, indicating that $x_1$ is not different between treatment and control towns. On the contrary, $x_2$ is significantly different between the two groups. This is not necessarily a concern for the DiD approach, since $x_1$ and $x_2$ are controls, and therefore should not directly effect our estimates for the treatment effect. What matters is that x1 and x2 are independent from the treatment assignment, and are not affected by the treatment. It doesn’t affect the validity of the DiD approach as long as we control for it and as long a the parallel trend assumption holds. The coefficient is almost the same, controlling for covariates only increased precision.

<!-- Randomized controlled trial ensures that allocation of towns to treatments is left purely to chance. The characteristics of towns that may influence outcome are distributed between treatment groups so that any difference in outcome can be assumed to be due to the intervention. However, imbalance between groups in baseline variables that may influence outcome can bias statistical tests, a property sometimes referred to as chance bias. Observed differences in outcome between groups in a particular trial could by chance be due to characteristics of the towns, not treatments.  -->

<!-- Some protection against chance bias is given by stratified randomisation or minimisation and by adjusting in the statistical analysis for baseline variables. -->


```{r, results='asis', message=F, warning=F}
did.ols1 <- plm(y ~ after*treatment + x1 + x2, data=d.did, 
                effect="individual", model="pooling", index=c("town", "t"))
vcov.ols1 <- vcovHC(did.ols1, type="sss", cluster="group")
se.ols1 <- sqrt(diag(vcov.ols1))
stargazer(did.ols, did.ols1, se=list(se.ols, se.ols1), single.row=T, header=F, 
          keep.stat=c("n", "rsq"), title="OLS with SE clustered by town")
```

In table 3, we see a decrease both in the coefficient of interest and clustered standard error from column (1) to column (2) but the difference between the coefficients of interest is trivial. After controlling for covariates, the standard error would be smaller (stronger statistical power). 

## 3.4

```{r, results='asis', message=F, warning=F}
did.fe <- plm(y ~ x1 + x2 + after*treatment, data=d.did, 
              effect="twoways", model="within", index=c("town", "t"))
vcov.fe <- vcovHC(did.fe, type="sss", cluster="group")
se.fe <- sqrt(diag(vcov.fe))
stargazer(did.ols, did.ols1, did.fe, se=list(se.ols, se.ols1, se.fe), 
          single.row=T, keep.stat=c("n", "rsq"), header=F, 
          title="FE regression with SE clustered by town")
```

The coefficient of interest from FE regression is similar to the one from OLS (without control variables) and OLS (with control variables) because we use clustered standard errors in all regression. Fixed-effects estimation takes into account unobserved unit-sepcific and time-specific confounders at the same time while cluster-adjusted standard errors account for within-cluster correlation or heteroscedasticity which the fixed-effects estimator does not take into account.

<!-- Cluster-adjusted standard error account for within-cluster correlation or heteroscedasticity which the fixed-effects estimator does not take into account -->



## 3.5

```{r}
ate <- rep(0, 10)
se <- rep(0, 10)
for (i in 1:10) {
  indicator <- d.did$t == i
  lm.fit <- lm(y ~ treatment, data=d.did, subset=indicator)
  ate[i] <- coef(summary(lm.fit))[2, 1]
  se[i] <- coef(summary(lm.fit))[2, 2]
}
se.bands <- cbind(ate+2*se, ate-2*se) # 95% confidence interval
```

```{r}
plot(1:10, ate, xlab="Time", ylab="Coefficients")
lines(ate, lwd=2, col="blue")
abline(v=5, col="lightblue", lwd=2)
matlines(1:10, se.bands, lwd=1, col="blue", lty=3)
title(main="Coefficients vs. Time")
```

```{r}
d.plot <- d.did %>% 
  group_by(treatment, t) %>% 
  summarise(y=mean(y), .groups="drop")
d.plot$treatment <- factor(d.plot$treatment)
```

```{r}
ggplot(d.plot, aes(x=t, y=y), color=treatment) + 
  geom_line(aes(color=treatment)) + 
  geom_point(aes(color=treatment)) + 
  geom_vline(xintercept=5) + 
  scale_color_manual(values=c("red", "blue")) + 
  theme_minimal() + ggtitle("Outcome vs. Time")
```

The parallel pre-trend assumption does not seem to hold.

## 3.6

```{r}
sub.did <- d.did[d.did$t %in% 1:5, ]
sub.did$placebo.after <- ifelse(sub.did$t %in% 4:5, 1, 0)
```

```{r, results='asis', message=F, warning=F}
did.placebo <- lm(y ~ placebo.after*treatment, data=sub.did)
did.placebo1 <- plm(y ~ placebo.after*treatment, data=sub.did,
                   effect="individual", model="pooling", index=c("town", "t"))
stargazer(did.placebo, single.row=T, header=F, keep.stat=c("rsq", "n"), 
          title="Placebo Test")
```

The placebo test is performed using a fake treatment group. The fake treatment group is constructed before the program. We should expect insignificant impact in order to support parallel pre-trend assumption. In table 5, however, we see that the coefficient of interest (interaction term) is statistically different from zero. The placebo test reveals significant impact. Therefore, the parallel pre-trend assumption is not supported.

## 3.7

\textcolor{cyan}{This suggests that the method used in the first two subsections gave us biased coefficient as we do not have the parallel trends assumption holding for all time periods before the treatment hits.}
